---
title: "Model Adequacy Checking - Examples from Montgomery/Peck/Vining"
author: "Jeremy Buss"
date: '2021-10-23'
slug: []
categories:
- Introduction to Linear Regression Analysis - Montgomery/Peck/Vining
- Applied Stats
- Regression Analysis
tags:
- R
- linear regression
draft: false
katex: yes
---
```{r echo=FALSE}
library(blogdown)
```

For the example data we will leverage the CRAN package "MPV".

```{r message=FALSE}
#install.packages("MPV")
library(MPV)
```


# Example 4.1: The Delivery Time Data

Let's define the dataframe for the SoftDrink data and take a look at the first 5 rows:

```{r}
SoftDrink <- data.frame(MPV::softdrink)
colnames(SoftDrink) <- c("DeliveryTime", "NumberCases", "Distance")
head(SoftDrink, 5)
```

With the data ready we will create:

* **SoftDrink.lm**: multivariate linear model
* **SoftDrink.lm.sum**: the summary of the linear model 

```{r}
SoftDrink.lm <- lm(DeliveryTime ~ NumberCases + Distance, data=SoftDrink)
SoftDrink.lm.sum <- summary(SoftDrink.lm)
```

From the linear model we see that our linear model equation is: 

DeliveryTime = `r SoftDrink.lm$coefficients[1]` + `r SoftDrink.lm$coefficients[2]` x #ofCases + `r SoftDrink.lm$coefficients[3]` x Distance

Now that we have the linear model let's figure out some of the residuals:

## Residual ("vanilla")

The basic, run o' the mill residuals 

$$
e_i = y_i - \hat{y}_i
$$

These are available directly from the linear regression model:

```{r}
SoftDrink.lm.residuals <- SoftDrink.lm$residuals
head(SoftDrink.lm.residuals,5)
```

## Standardized Residuals

For the standardized Residuals we use the basic residuals and divide by the square root of Mean Square Residual

$$
d_i = \frac{e_i}{\sqrt{MS_{Res}}}
$$

Within R we can grab the sigma value from the summary of the linear model and leverage the residuals that we defined in the previous step:

```{r}
SoftDrink.lm.MS_res <- SoftDrink.lm.sum$sigma^2
SoftDrink.lm.StandardizedResiduals <- SoftDrink.lm.residuals / sqrt(SoftDrink.lm.MS_res)
head(SoftDrink.lm.StandardizedResiduals,5)
```

## Studentized Residuals 

For the **studentized residuals** we need to adjust the denominator of the residual to have constant variance regardless of location of $\mathbf{x_i}$.

This gives us:

$$
r_i = \frac{e_i}{\sqrt{MS_{res}(1-h_{ii})}}
$$
We would generally need to determine the values for $h_{ii}$ to calculate this residual. The $h_{ii}$ values can be determined with the following:

```{r}
SoftDrink.lm.HatValues <- lm.influence(SoftDrink.lm)$hat
head(SoftDrink.lm.HatValues,5)
```

However we are able to leverage a function to determined the studentized residuals straight away:

```{r}
SoftDrink.lm.StudentizedResiduals <- rstandard(SoftDrink.lm)
head(SoftDrink.lm.StudentizedResiduals,5)
```

## PRESS Residuals

The next flavor of residual is the **PRESS** residual which is the ordinary residual weighted by the diagonal elements of the hat matrix $h_{ii}$ it can be espressed as:

$$
e_{(i)}=y_i - \hat{y}_{(i)}
$$
Additionally we can express it as:

$$
e_{(i)} = \frac{e_i}{1-h_{ii}}, \space i=1,2,\dots,n
$$

This last definition is what we will use to calculate the values.

```{r}
SoftDrink.lm.PRESSResiduals <- SoftDrink.lm.residuals / (1 - SoftDrink.lm.HatValues)
head(SoftDrink.lm.PRESSResiduals,5)
```

## R-Student Residuals

With the last residual, R-Student Residuals, we use $S_{(i)}^2$ as an estimate of $\sigma^2$ based on the data with the $i$-th observation removed:

$$
S_{(i)}^2 = \frac{(n-p)MS_{res}-e_i^2/(1-h_{ii})}{n-p-1}
$$

and placing this estimate into the following:

$$
t_i = \frac{e_i}{\sqrt{S_{(i)}^2(1-h_{ii})}}
$$

However there is a package which makes it much easier:

```{r}
library(MASS)
SoftDrink.lm.RStudentResiduals <- studres(SoftDrink.lm)
head(SoftDrink.lm.RStudentResiduals,5)
```

## Combined Residuals Dataframe

```{r}
SoftDrink.Residual_df <- data.frame(
    SoftDrink.lm.residuals,
    SoftDrink.lm.StandardizedResiduals,
    SoftDrink.lm.StudentizedResiduals,
    SoftDrink.lm.HatValues,
    SoftDrink.lm.PRESSResiduals,
    SoftDrink.lm.RStudentResiduals
)

colnames(SoftDrink.Residual_df) <- c(
    "Residuals $e_i$",
    "Standardized Residuals $d_i$",
    "Studentized Residuals $r_i$", 
    "Hat Matrix Diagonals $h_{ii}$", 
    "PRESS Residuals $e_{(i)}$", 
    "R-Student Residuals $t_i$"
)

library(knitr)

kable(SoftDrink.Residual_df, caption="Soft Drink Linear Model Residuals")
```


# 4.2 Plot of Residuals against the Fitted Values $\hat{y_i}$

We use a type of plot called [QQ-plot](https://en.wikipedia.org/wiki/Q%E2%80%93Q_plot). This "...is a graphical method for comparing two probability distributions by plotting their quantiles against each other"(Wikipedia)

```{r DeliveryTimeQQPlot, fig.cap="QQ Plot for Soft Drink Model"}

#dev.new(width=100, height=100, unit="px")
qqnorm(SoftDrink.lm.RStudentResiduals,
       ylab="R Student Residuals",
       xlab="Normal Scores",
       main="Delivery Time"
       , datax=TRUE
)
qqline(SoftDrink.lm.RStudentResiduals, distribution=qnorm, datax=TRUE)
```

# 4.3 Plot of Residuals against Fitted Value

```{r}
plot(
    fitted(SoftDrink.lm), 
    SoftDrink.lm.RStudentResiduals,
    xlab="Fitted Values",
    ylab="Externally Studentized Residual",
    main="Versus Fits"
)
abline(h=0)
```

# 4.4 Residuals against Regressors

```{r}

plot(
    SoftDrink$NumberCases,
    SoftDrink.lm.RStudentResiduals,
    xlab="Number of Cases",
    ylab="Externally Studentized Residual",
    main="Residuals Versus Number of Cases"
)
abline(h=0)
```
```{r}
plot(
    SoftDrink$Distance, 
    SoftDrink.lm.RStudentResiduals,
    xlab="Distance",
    ylab="Externally Studentized Residual",
    main="Residuals Versus Distance"
)
abline(h=0)
```


# 4.5 Partial Regression Plots

```{r}
SoftDrink.PrtlReg.Delivery.Distance.y.lm <- lm(DeliveryTime ~ Distance, data=SoftDrink)
SoftDrink.PrtlReg.Delivery.Distance.y.lm.Residuals <- resid(SoftDrink.PrtlReg.Delivery.Distance.y.lm)
SoftDrink.PrtlReg.Delivery.Distance.x.lm <- lm(NumberCases ~ Distance, data=SoftDrink)
SoftDrink.PrtlReg.Delivery.Distance.x.lm.Residuals <- resid(SoftDrink.PrtlReg.Delivery.Distance.x.lm)

plot(
    SoftDrink.PrtlReg.Delivery.Distance.x.lm.Residuals, 
    SoftDrink.PrtlReg.Delivery.Distance.y.lm.Residuals,
    xlab="Cases",
    ylab="Time",
    main="Partial Regression Plot - Distance"
)
```

```{r}
SoftDrink.PrtlReg.Delivery.NumberCases.y.lm <- lm(DeliveryTime ~ NumberCases, data=SoftDrink)
SoftDrink.PrtlReg.Delivery.NumberCases.y.lm.Residuals <- resid(SoftDrink.PrtlReg.Delivery.NumberCases.y.lm)
SoftDrink.PrtlReg.Delivery.NumberCases.x.lm <- lm(Distance ~ NumberCases, data=SoftDrink)
SoftDrink.PrtlReg.Delivery.NumberCases.x.lm.Residuals <- resid(SoftDrink.PrtlReg.Delivery.NumberCases.x.lm)

plot(
    SoftDrink.PrtlReg.Delivery.NumberCases.x.lm.Residuals, 
    SoftDrink.PrtlReg.Delivery.NumberCases.y.lm.Residuals,
    xlab="Cases",
    ylab="Time",
    main="Partial Regression Plot - Number Cases"
)
```

# 4.6 Delivery Time Data - Residuals

We add a new residual to our chart of residuals called the PRESS statistic 


```{r}
SoftDrink.Residual_df <- data.frame(
    SoftDrink.lm.residuals,
    SoftDrink.lm.StandardizedResiduals,
    SoftDrink.lm.StudentizedResiduals,
    SoftDrink.lm.HatValues,
    SoftDrink.lm.PRESSResiduals,
    SoftDrink.lm.RStudentResiduals,
    SoftDrink.lm.PRESSResiduals^2
)

colnames(SoftDrink.Residual_df) <- c(
    "Residuals $e_i$",
    "Standardized Residuals $d_i$",
    "Studentized Residuals $r_i$", 
    "Hat Matrix Diagonals $h_{ii}$", 
    "PRESS Residuals $e_{(i)}$", 
    "R-Student Residuals $t_i$",
    "PRESS Statistic $e_{(i)}^2$"
)


kable(SoftDrink.Residual_df, caption="Soft Drink Linear Model Residuals")
```
